---
title: "Project 2"
author: "Lisa Siefker"
date: "2025-03-22"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr) 
library(tidyverse)
library(ggplot2) 
library(vcd)
```

```{r}
brant <- read.table(file="project4data.txt", col.names = c("observ", "altitude", "lateral", "flight"))
head(brant)
```

```{r}
# get dataframe dimensions and summary data 
print(dim(brant))
summary(brant)
```

```{r}
# check for missing values
sum(is.na(brant))
```

```{r}
# altitude is randomly assigned for each flight to one of nine discrete levels
# get counts of altitudes 
altitude_counts <- table(brant$altitude)
print(altitude_counts)
```

```{r}
# add column that classifies altitude into three categories: <3, 3 - 6, > 6
brant_1 <- brant
brant_1$altitude_cat <- cut(
  brant_1$altitude,
  breaks = c(-Inf, 3, 6, Inf),  
  labels = c("<3", "3-6", ">6")  
)

head(brant_1, 2)
```

```{r}
# get counts in each altitude category
table(brant_1$altitude_cat)
```

Note that altitude categories are unbalanced.

```{r}
# add column that classifies lateral distance into four classes: <10, 10-19, 20-29, >30 
brant_1$lateral_cat <- cut(
  brant_1$lateral,
  breaks = c(-Inf, 10, 20, 30, Inf),
  labels = c("<10", "10-19", "20-29", ">30")  
)

head(brant_1, 2)
```

```{r}
# get counts in each lateral category
table(brant_1$lateral_cat)
```

Note that lateral categories are unbalanced.

```{r}
# get counts of flight response (1=yes, 0=no)
flight_counts <- table(brant_1$flight)
print(flight_counts)
```

```{r}
# scatterplot altitude v. lateral 
ggplot(brant_1, aes(x = lateral, y = altitude, color = factor(flight))) +
  geom_jitter(width = 0.2, height = 0.2) +  # Add jitter to avoid overplotting
  labs(
    title = "Scatterplot of Altitude vs Lateral by Flight Response",
    x = "Lateral",
    y = "Altitude",
    color = "Flight Response"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("0" = "lightgreen", "1" = "darkblue")) 
```

The scatter plot shows that smaller lateral distance and higher altitudes tend to result in a flight response.

Q1: What is the effect of helicopter altitude on the flight response of Pacific brant?

```{r}
# Bar plot of flight response by altitude category
barplot(table(brant_1$altitude_cat, brant_1$flight), 
        main = "Flight Response by Altitude Category", 
        xlab = "Flight Response", 
        ylab = "Count", 
        col = c("lightblue", "lightgreen", "forestgreen"), 
        legend = rownames(table(brant_1$altitude_cat, brant_1$flight))) 
```

This plot shows that overall there were more flight responses than non-responses in the dataset. Both flight responses and non-responses occurred at every altitude category. At the \>6 altitude category, the plot indicates that significantly more flights had a response than did not have a response.

```{r}
# Grouped bar plot
ggplot(brant_1, aes(x = altitude_cat, fill = factor(flight))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Flight Response by Altitude Category",
    x = "Altitude Category",
    y = "Count",
    fill = "Flight Response"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "darkblue"))

```

This is another way to visualize flight response and non-response by altitude category. It's clear that more flocks had a response than did not have a response at every altitude category.

Does flight response have a different mean altitude or lateral distance? This question is to determine the difference in the calculated means of our response vs. non-response group without considering categories of altitude or lateral distance.

The mean for altitude is lower in the non-response group, while the mean for lateral distance is significantly smaller for the response group. To verify these differences are significant, t-tests were run and analyzed and then a Wilcoxon sign test.

```{r}
# find the means
mean((brant_1$altitude[brant_1$flight == 1]))
mean((brant_1$altitude[brant_1$flight == 0]))

mean((brant_1$lateral[brant_1$flight == 1]))
mean((brant_1$lateral[brant_1$flight == 0]))

# run t-tests on lateral and altitude
t.test(brant_1$altitude ~ brant_1$flight)
t.test(brant_1$lateral ~ brant_1$flight)

#test assumptions of t-tests
lateral_flight_0 <- brant_1$lateral[brant_1$flight == 0]
lateral_flight_1 <- brant_1$lateral[brant_1$flight == 1]
altitude_flight_0<-brant_1$altitude[brant_1$flight==0]
altitude_flight_1<-brant_1$altitude[brant_1$flight==1]

qqnorm(lateral_flight_0,main="Distance: No Flight Response")
qqline(lateral_flight_0)
qqnorm(lateral_flight_1,main="Distance: Flight Response")
qqline(lateral_flight_1)
qqnorm(altitude_flight_0,main="Altitude: No Flight Response")
qqline(altitude_flight_0)
qqnorm(altitude_flight_1,main="Altitude: Flight Response")
qqline(altitude_flight_1)
```

While the t-tests show there is overwhelming evidence that the mean lateral distance is much lower for the response group than for the non-response group (p value of 2.2e-16) and very strong evidence that the mean altitude distance is lower for the non-response group (p value of .007589), the data may not meet all the assumptions required to use t-tests. The Q-Q plots show that our data here is most likley not Normally distributed and we may need to consider nonparametic tests.

```{r}
# run Wilcoxon sign test
wilcox.test(lateral~flight, data = brant_1)
wilcox.test(altitude~flight, data = brant_1)
```

Based on the Wilcoxon sign test, there is overwhelming evidence that mean lateral distance between the flight and flight response groups are different. The altitude also show some evidence. With a p-value of .04 we would reject the null hypothesis (that mean altitude between the two groups is the same) with a reasonable degree of confidence.

```{r}
#plotting the lateral effect 
barplot(table(brant_1$lateral_cat, brant_1$flight), 
        main = "Flight Response by Lateral Category", 
        xlab = "Flight Response", 
        ylab = "Count", 
        col = c("lightblue", "lightgreen", "forestgreen", "darkorchid1"), 
        legend = rownames(table(brant_1$lateral_cat, brant_1$flight))) 
```

This plot shows that there were significantly more flight responses at the closest lateral distance. While non-responses were spread across the data in each bracket of distance, by far the greatest number of responses occurred in the closest bracket.

```{r}
# Grouped bar plot
ggplot(brant_1, aes(x = lateral_cat, fill = factor(flight))) +
  geom_bar(position = "dodge") +
  labs(
    title = "Flight Response by Lateral Category",
    x = "Lateral Category",
    y = "Count",
    fill = "Flight Response"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "darkblue"))
```

This plot shows an overwhelmingly higher number of close (\<1000 m) helicopter flights causing a flight response. In all other distances, there were more non-responses than responses, with a significant drop off proportionally once the helicopters were more than 2000 m away.

```{r}
# use chi-squared test to check for significant association between altitude_cat and flight 
# tests whether the distribution of flight responses differs across all levels of altitude category
chisq.test(table(brant_1$altitude_cat, brant_1$flight))
```

Interpretation of the Chi-Square Test:

Null Hypothesis: There is no association between altitude category and flight response Alternative Hypothesis: There is an association between altitude category and flight response

The P-value is less than the significance level of 0.05, so we reject the null hypothesis and conclude that there is a statistically significant association between altitude category and flight response.

```{r}
#Use logistic regression to model the probability of a flight response as a function of altitude category

# Convert altitude_cat to factor 
brant_1$altitude_cat <- as.factor(brant_1$altitude_cat)

# Fit a logistic regression model predicting flight response by altitude category
logistic_model <- glm(flight ~ altitude_cat, data = brant_1, family = binomial)

# Summarize the model
print(summary(logistic_model))

```

Interpreting Model Output:

The logistic regression model predicts the probability of a flight response based on the altitude category.

Null hypothesis: Altitude category has no effect on flight response Alternative hypothesis: At least one altitude category has an effect on flight response

The p-value for altitude category \>6 is 0.0009. The pvalue is less than the significance level of 0.05, so we reject the null hypothesis and conclude that helicopters flying at a level of \>6 have a significant impact on flight response at the 5% level of significance.

The p-values of the \<3 and 3-6 altitude categories are greater than 0.05, indicating that helicopters flying at those altitudes do not have a significant effect on flight response.

```{r}
# Convert variables to factor
brant_1$altitude_cat <- as.factor(brant_1$altitude_cat)
brant_1$lateral_cat <- as.factor(brant_1$lateral_cat)

# Fit the logistic regression model
logistic_model_1 <- glm(flight ~ altitude_cat + lateral_cat, data = brant_1, family = binomial)

# Summarize the model
summary(logistic_model_1)

```

Altitude category 3-6 is the only variable category that does not have a significant effect on flight, so I will remove it from the model.

```{r}
# Filter the data to exclude altitude_cat3-6
brant_filtered <- brant_1[brant_1$altitude_cat != "3-6", ]

# Fit the logistic regression model on the filtered data
logistic_model_2 <- glm(flight ~ altitude_cat + lateral_cat, data = brant_filtered, family = binomial)

# Summarize the model
summary(logistic_model_2)
```

Lateral category \>30 does not have a significant effect on flight, so I will remove it from the model.

```{r}
# Filter the data to exclude lateral_cat>30
brant_filtered_2 <- brant_filtered[brant_filtered$lateral_cat != ">30", ]

# Fit the logistic regression model on the filtered data
logistic_model_3 <- glm(flight ~ altitude_cat + lateral_cat, data = brant_filtered_2, family = binomial)

# Summarize the model
summary(logistic_model_3)
```

All remaining variable categories appear to have a significant impact on flight response.

```{r}
# Filter the data to exclude lateral_cat>30
brant_filtered_2 <- brant_filtered[brant_filtered$lateral_cat != ">30", ]

# Fit the logistic regression model on the filtered data and add an interaction between lat and altitude
logistic_model_4 <- glm(flight ~altitude_cat+ lateral_cat+altitude_cat*lateral_cat, data = brant_filtered_2, family = binomial(link=logit))

# Summarize the model
summary(logistic_model_4)
```

```{r}
# Interaction plot
interaction.plot(brant_1$altitude_cat, brant_1$lateral_cat, brant_1$flight, xlab = 
                   "Altitude bracket", ylab="Mean of Flight Response", trace.label="Lateral distance bracket")

# alternate format

intersum <-brant_1 %>% group_by(altitude_cat, lateral_cat) %>% summarize (prop_flight_1 = mean(flight, na.rm=TRUE), n = n())
ggplot(intersum, aes(x = altitude_cat, y = prop_flight_1, color = lateral_cat, group = lateral_cat)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Proportion of Flight by Altitude and Lateral Category",
       x = "Altitude Category",
       y = "Proportion (Flight = 1)",
       color = "Lateral Category") +
  ylim(0, 1)

```

If we look at the plotted proportions, it appears that all altitudes in the shortest lateral distance have a high proportion of flight responses and the line is relatively flat (that is, at that close distance, there was not a significant interaction between altitude and lateral distance). In the mid-range altitude bracket (300-600m), there appears to be an interaction with the two highest brackets for lateral distance. That is, when the helicopters were more than 3000 m laterally away but at a height of 300-600 meters, there was a higher proportion of flight responses than when the helicopters were within 2000-2900 m.

```{r Confidence Interval, paged.print=TRUE}

confint(logistic_model_4)


```
```{r}
exp(-3.765)
```
```{r}
glm(formula = flight ~ altitude_cat + lateral_cat + altitude_cat * 
    lateral_cat, family = binomial(link = logit), data = brant_1)
```
```{r}
# Odds Ratios
exp(-3.7651)  # OR for lateral 10-19 at altitude≤6
exp(-5.1240)  # OR for lateral 20-29 at altitude≤6
exp(-3.7651 + 3.1650)  # OR for lateral 10-19 at altitude>6
exp(-5.1240 + 3.2022)  # OR for lateral 20-29 at altitude>6

# Probabilities
plogis(1.9459)                  # Reference probability
plogis(1.9459 - 3.7651)         # Lateral 10-19 at altitude≤6
plogis(1.9459 - 5.1240)         # Lateral 20-29 at altitude≤6
plogis(1.9459 - 3.7651 + 3.1650) # Lateral 10-19 at altitude>6
```


